Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-03-04 paper/paper.tex
pass on boxplots
done with section 2

This paper focuses on rule-based \nids{}, which are very popular in
industry~\cite{proofpoint-etpro,snort-rule-subscriptions}. A
\emph{rule-based intrusion detector}\footnote{\aka\ signature-based
  intrusion detector.} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:pingscan-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. This
particular rule prescribes a method to capture an Active
Reconnaissance attack for discovering IPS of hosts in a network (see
Section~\ref{sec:active-recon}). Rulesets exist for Snort~\cite{snort}
and Suricata~\cite{suricata}, providing protection against existing
attacks. Unfortunately, attackers actively look for new mechanisms to
exploit systems. For example,
Figure~\ref{fig:distribution-rules-per-month} shows the number of
rules added per month in the year of 2019 to the ``Emerging Threats
Open Ruleset'', a popular public data set of \suri\ rules. We used the
``creation-date'' field associated with each entry in the ruleset. The
data provides some evidence that new rules are created often.

@@

Figure~\ref{fig:pingscan-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.  A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The \emph{action} appears as the
first word in the rule description (an \CodeIn{alert}, in this
case). An action denotes the task that needs to be executed if the
rule pattern is satisfied.  In this example, an alert message will be
sent to system administrators if the rule pattern is
satisfied\footnote{Technically, an IPS (Intrusion Prevention System)
  differs from an IDS in the action taken when an intrusion is
  observed. IPS systems block malicous traffic whereas IDS system only
  alert them. However, IDS and IPS do not differ in the approaches
  used to detect the intrusion~\cite{ids-ips}.\Mar{check}}. The
\emph{header} comes after the action in the rule description. It
restricts the information flow covered by the rule. The header of this
rule is \CodeIn{icmp any any -> any any}. It instructs Suricata to
inspect \CodeIn{icmp} traffic flowing from any machine and port to any
machine and port.\MyComment{ The variables \CodeIn{\$HOME\_NET} and
  \CodeIn{\$EXTERNAL\_NET} are configurable.} Finally, \emph{rule
  options} come after the header. It is a semi-colon-separated
sequence of key-value pairs. Rule options serve to document the
analyzed traffic (\eg{}, ``msg'', ``classtype'', and ``sid'') and to
characterize the attack pattern (\eg, ``flags'', ``threshold'', and
``content''). \tname{} produces options of the second kind, which are
\emph{relevant} to capture the attack. For example, the option
\CodeIn{content:<arg>} checks if the argument is present in the
payload of the message.

@@

Table~\ref{table:rules} shows a sample of non-documentation options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column ``Description'' summarizes
purpose. The option \CodeIn{content} is very common in rules to
capture http attacks, which are the most common kind of rules. This
option takes a sequence of bytes as argument (\eg{},
\CodeIn{content:''SELECT ''}) and checks if that sequence is present
in the payload of the message. Detailed description of the Suricata
rules can be found elsewhere~\cite{suri-rule-format}. It is worth
noting that Suricata supports a total of \totoptions\ distinct options
covering \numproto\ distinct protocols.

@@

Rule options are often associated with specific fields of network
packets. Let's consider the ICMP protocol as an
example. Figure~\ref{fig:icmp-packet-layout} shows the layout of an
ICMP packet. The packet fields ``Flags'' and ``Time to Live'',
declared in the IP header, are mapped to the options \CodeIn{flagbits}
and \CodeIn{ttl}, respectively. Likewise, the fields ``Type'' and
``Code'', declared in the ICMP header, are mapped to the options
\CodeIn{itype} and \CodeIn{icode}. The values of the options ``ttl'',
``itype'', and ``icode'' are the same as the value of the
corresponding fields in the ICMP packet. For the field ``Flags'' the
value of the corresponding option is encoded in a bitvector. For
example, if the ``Flags'' field stores the value 1010, the
corresponding rule option will be \CodeIn{flagbits:RM}, as each
position of the vector encodes a different character in the
\CodeIn{flagbits} rule option. The value of several options can be
determined through transformation functions, as those described
above. One exception to this is the option \CodeIn{threshold}, which
is typically used in denial-of-service attacks. That option operates
over sequence of messages as opposed to a single
message. Consequently, it is not possible to determined its values
from a single message.

@@

          every node near coord/.append style={font=\tiny},

@@

@@

\begin{figure}[t!]
\centering
\includegraphics[scale=0.4]{R/scripttag/scripttag.pdf}
\caption{Impact of number of positive examples.}
\label{fig:impact-number-attacks}
\end{figure}

>>>>

This paper focuses on rule-based \nids{}, which are very popular in
industry~\cite{proofpoint-etpro,snort-rule-subscriptions}. A
\emph{rule-based intrusion detector}\footnote{\aka\ signature-based
  intrusion detector.} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:pingscan-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. This
particular rule prescribes a method to capture an Active
Reconnaissance attack for discovering IPS of hosts in a network (see
Section~\ref{sec:active-recon}). Rulesets exist for Snort~\cite{snort}
and Suricata~\cite{suricata}, providing protection against existing
attacks. Unfortunately, attackers actively look for new mechanisms to
exploit systems. For example,
Figure~\ref{fig:distribution-rules-per-month} shows the number of
rules added per month in the year of 2019 to the ``Emerging Threats
Open Ruleset''~\cite{emerging-threats-open}, a popular public data set
of \suri\ rules. We used the ``creation-date'' field associated with
each entry in the ruleset. The data provides some evidence that new
rules are created often.

@@

Figure~\ref{fig:pingscan-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}.  A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The \emph{action} appears as the
first word in the rule description (an \CodeIn{alert}, in this
case). An action denotes the task that needs to be executed if the
rule pattern is satisfied.  In this example, an alert message will be
sent to system administrators if the rule pattern is
satisfied\footnote{Technically, an IPS (Intrusion Prevention System)
  differs from an IDS in the action taken when an intrusion is
  observed. IPS systems block malicous traffic whereas IDS system only
  alert them. However, IDS and IPS do not differ in the approaches
  used to detect the intrusion~\cite{ids-ips}.\Mar{check}}. The
\emph{header} comes after the action in the rule description. It
restricts the information flow covered by the rule. The header of this
rule is \CodeIn{icmp any any -> any any}. It instructs Suricata to
inspect \CodeIn{icmp} traffic flowing from any machine and port to any
machine and port.\MyComment{ The variables \CodeIn{\$HOME\_NET} and
  \CodeIn{\$EXTERNAL\_NET} are configurable.} Finally, \emph{rule
  options} come after the header. It is a semi-colon-separated
sequence of key-value pairs. Rule options serve to document the
analyzed traffic (\eg{}, ``msg'', ``classtype'', and ``sid'') and to
characterize the attack pattern (\eg, ``flags'', ``threshold'', and
``content''). \tname{} produces options of the second kind, which are
\emph{relevant} to capture the attack. For example, the option
\CodeIn{content:<arg>} checks if the argument is present in the
payload of a message.

@@

Table~\ref{table:rules} shows a sample of non-documentation options of
Suricata. Column ``\#'' shows the option number, column ``Name'' shows
the name of the option, and column ``Description'' summarizes
purpose. The option \CodeIn{content} is common in rules to capture
http attacks and http rules are the most common in the Emerging
Threats Open Ruleset~\cite{emerging-threats-open}. The content option
takes a sequence of bytes as argument and checks if that sequence is
present in the payload of the message.  For example, the option
\CodeIn{content:''SELECT''} checks if the string ``SELECT'' is present
in the payload.  Detailed description of the Suricata rules can be
found elsewhere~\cite{suri-rule-format}. It is worth noting that
Suricata supports a total of \totoptions\ distinct options covering
\numproto\ distinct network protocols.

@@

We mentioned above that http rules often used content options, which
associate with the message payload. For non-http rules, options are
often associated with packet fields. Let us consider a rule for an
ICMP attack as example. Figure~\ref{fig:icmp-packet-layout} shows the
layout of an ICMP packet. The packet fields ``Flags'' and ``Time to
Live'', declared in the IP header, are mapped to the \suri\ options
\CodeIn{flagbits} and \CodeIn{ttl}, respectively. Likewise, the fields
``Type'' and ``Code'', declared in the ICMP header, are mapped to the
options \CodeIn{itype} and \CodeIn{icode}, respectively. The values of
the options ``ttl'', ``itype'', and ``icode'' are the same as the
value of the corresponding fields in the ICMP packet. For the field
``Flags'' the value of the corresponding option is encoded in a
bitvector. For example, if the ``Flags'' field stores the value 1010,
the corresponding rule option will be \CodeIn{flagbits:RM}, as each
position of the vector encodes a different character in the
\CodeIn{flagbits} rule option. In summary, the value of several
options can be determined through transformation functions, as those
informally described above. One exception to this is the option
\CodeIn{threshold}, which is typically used in denial-of-service
attacks. That option operates over sequence of messages as opposed to
a single message. Consequently, it is not possible to determine its
values from a single message.

@@

          xticklabel style={font=\footnotesize},          

@@

          xticklabel style={font=\footnotesize},          

@@

\begin{figure}[t!]
  \centering
  \begin{subfigure}{.2\textwidth}
    \centering
    \includegraphics[scale=0.4,trim=20 0 130 0,clip]{R/idqaccess/idqaccess.pdf}
    \caption{idqaccess}
  \end{subfigure}%
  \begin{subfigure}{.2\textwidth}
    \centering
    \includegraphics[scale=0.4,trim=15 0 130 0,clip]{R/scripttag/scripttag.pdf}
    \caption{scripttag}    
  \end{subfigure}%
\caption{Impact of number of positive examples.}
\label{fig:impact-number-attacks}
\end{figure}
